\sys: Light-touch Data-Parallel Shell Processing

### 2020-10-05 paper/draft.tex
Pass over 4

@@

\sys's core is an abstract dataflow graph (DFG) model that is used as the intermediate representation on which \sys performs optimizations.
%% representation that is expressive enough to model data parallelism in shell programs.
%% support a large subset of shell programs.\nv{Immediate Q: How Large? Which programs are not supported? Why?}
\sys first lifts sections of the input script to the DFG
representation, it performs transformations to expose parallelism, and
then instantiates each DFG back to a parallel shell script.  A
fundamental difference with other DFG models is that \sys's DFG encodes the
order in which a node reads its inputs, which in turn enables a set of
graph transformations that can be iteratively applied to expose
parallelization opportunities for \tsta and \tpur commands.
%% data parallelism as task parallelism.\nv{What does this mean? What if they are fundamentally different in the reivewer's mind? Example?}
%% These transformations can then be applied in an iterative way, optimizing the program to utilize available computational resources. 
In \Cref{graph-components} we define the DFG model and describe the
correspondence with shell scripts, and in \Cref{ir:transformations} we
describe the graph transformations that \sys performs.

\subsection{Parallelizable Regions}
\label{dataflow-regions}
The dataflow model and associated transformations are applied only to parallelizable regions.
% 
% To optimize a script, \sys translates it to a DFG~\sx{ir}.
% As there are several program fragments that cannot be parallelized---\eg commands connected with \ttt{\&\&}, ~\S\ref{bg}---\sys introduces the notion of \emph{parallelizable program regions} and a translation pass for converting them to DFGs.
% 
%% In addition, the shell's highly dynamic
%% nature compounds the challenge---\eg by not knowing what files does an
%% unexpanded command argument refer to.

%% \kk{Give examples of writing to environment variables, producing
%%   irreversible side effects, the ; and \&\& operators.}

%% To address the first issue, we introduce the notion of parallelizable
%% program regions and a translation pass for converting them to a
%% dataflow graph. To address the second issue, we limit dynamic
%% components by deferring translation as late as possible using a
%% wrapping scheme.

%% The translation pass identifies distributable regions and
%% translates each of them to a dataflow graph, provided the abstract
%% syntax tree (AST) of a shell script.  The AST is produced by
%% \ttt{LibDash}, a POSIX compliant parser~\cite{libdash}.

%
%% \kk{The goal of this subsection is to explain that Dish does a safe
%%   but effective analysis to identify which parts of the scripts to
%%   parallelize.}
%
Parallelizable regions are program sub-expressions that \sys deems may be parallelizable safely, \ie without breaking the program, based on the shell constructs they use.
%% respect to their sequential execution.
The search for these regions is guided by the shell language and the structure of a particular program.
These contain information about two types of components:
  fragments that can be executed independently and barriers that are natural synchronization points. % for enforcing synchronization.
Consider this fragment:

@@

\noindent
The \ttt{cat} and \ttt{grep} commands execute independently (and concurrently) in the standard shell, but \ttt{sort} waits for their completion prior to start.
Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus parallelizable regions, each not extending beyond \ttt{\&\&}.
% ---more precisely, they are maximal.

Intuitively, parallelizable regions correspond to sub-expressions of
the program that would be allowed to execute independently by
different processes in the POSIX standard~\cite{posix}. Larger
parallelizable regions can be composed from smaller ones using the
pipe operator (\ttt{|}) and the parallel-composition operator (\ttt{\&}).
Conversely, all other operators, such as the sequential composition operator (\ttt{;}) and the logical operators (\ttt{\&\&}, \ttt{||}), represent
barrier constructs that do not allow parallelizable regions to
permeate through.\nv{I will make this more precise---this paragraph is almost the same as overview}

@@

\heading{Translation Pass}
%
%% \kk{It might be a good idea to change this subsection based on the new
%%   as-late-as-possible architecture where everything is wrapped and
%%   translated at the final moment. Ideally, this would become the main
%%   point of this subsection, as the translation algorithm is kind of
%%   redundant when having the distributable regions too.}
%
\sys's front-end performs a depth-first search on the AST of
the given shell program.  During this pass, it extends the
parallelizable regions bottom-up, translating their
independent components to DFG nodes until a barrier
construct is reached. All subtrees not translatable to DFGs are kept as they are. The output of the translation pass is the
original AST where parallelizable regions have been replaced with
DFGs and calls to \sys's runtime primitives~\sx{optimizer}.

@@

\subsection{Definitions}
\label{graph-components}

@@

We first introduce basic notation used in this section.
For a set $D$, we write $D\kstar$ to denote the set of all finite words over $D$.
For words $x, y \in D\kstar$, we write $x \cdot y$ or $xy$ to denote their concatenation.
We write $\eps$ for the empty word.
We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there is a word $z$ such that $y = xz$.
The $\leq$ order is reflexive, antisymmetric, and transitive (\ie it is a partial order), and is often called the \emph{prefix order}.

@@

@@

After translating the parallelizable regions of the input script to DFGs and applying optimization passes~\sx{ir:transformations},
  \sys translates all DFGs back into a shell script.
Nodes of the graph are instantiated with the commands and flags they represent, and edges are instantiated as named pipes.
This section describes technical challenges related to execution of the resulting script and how they are addressed by \sys's custom primitives.
%% it instantiates the \todo{custom} split and relay nodes as highly
%% optimized commands to improve In this section we describe how \sys
%% achieves that, and several technical challenges that it addresses.

>>>>

\TODO{Add pointers}

@@

\sys's core is an abstract dataflow graph (DFG) model that is used as the intermediate representation on which \sys performs optimizations.
%% representation that is expressive enough to model data parallelism in shell programs.
%% support a large subset of shell programs.\nv{Immediate Q: How Large? Which programs are not supported? Why?}
\sys first lifts sections of the input script to the DFG
representation, it performs transformations to expose parallelism, and
then instantiates each DFG back to a parallel shell script.  A
fundamental difference with other DFG models is that \sys's DFG encodes the
order in which a node reads its input streams (not just the order of input elements per stream), which in turn enables a set of
graph transformations that can be iteratively applied to expose
parallelization opportunities for \tsta and \tpur commands.
%% data parallelism as task parallelism.\nv{What does this mean? What if they are fundamentally different in the reivewer's mind? Example?}
%% These transformations can then be applied in an iterative way, optimizing the program to utilize available computational resources. 
We start with the script-to-DFG tranlation~\sx{dataflow-regions}, 
  then define the DFG model and describe the correspondence with shell scripts~\sx{graph-components},
  discuss the transformations that \sys performs~\sx{ir:transformations}, and close with
  the DFG-to-script translation~\sx{backend}.

\subsection{Font-end: From Shell Scripts to DFGs}
\label{dataflow-regions}

% To optimize a script, \sys translates it to a DFG~\sx{ir}.
% As there are several program fragments that cannot be parallelized---\eg commands connected with \ttt{\&\&}, ~\S\ref{bg}---\sys introduces the notion of \emph{parallelizable program regions} and a translation pass for converting them to DFGs.
% 
%% In addition, the shell's highly dynamic
%% nature compounds the challenge---\eg by not knowing what files does an
%% unexpanded command argument refer to.
% 
%% \kk{Give examples of writing to environment variables, producing
%%   irreversible side effects, the ; and \&\& operators.}
%% To address the first issue, we introduce the notion of parallelizable
%% program regions and a translation pass for converting them to a
%% dataflow graph. To address the second issue, we limit dynamic
%% components by deferring translation as late as possible using a
%% wrapping scheme.
% 
%% The translation pass identifies distributable regions and
%% translates each of them to a dataflow graph, provided the abstract
%% syntax tree (AST) of a shell script.  The AST is produced by
%% \ttt{LibDash}, a POSIX compliant parser~\cite{libdash}.
% 
%
%% \kk{The goal of this subsection is to explain that Dish does a safe
%%   but effective analysis to identify which parts of the scripts to
%%   parallelize.}

\heading{Parallelizable Reions}
The DFG model and associated transformations are applied only to program sub-ex\-pres\-sions that \sys deems \emph{may} (but not must) be safely parallelizable---\ie ones that would not break sequential semantics, as long as the commands involved in these regions fall under the right classes.
%% respect to their sequential execution.
The search for these regions is guided by the shell language and the structure of a particular program.
These contain information about (i) fragments that can be executed independently, and (ii) barriers that are natural synchronization points. % for enforcing synchronization.
Consider this fragment:

@@

\noindent
The \ttt{cat} and \ttt{grep} commands execute independently (and concurrently) in the standard shell, but \ttt{sort} waits for their completion prior to start.
Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus parallelizable regions, each not extending beyond \ttt{\&\&}.\footnote{
  Can we make the commands \ttt{cmd1}, \ttt{cmd2}, \etc to ensure the readers understand this is driven by structure and not commands?}
% ---more precisely, they are maximal.

Intuitively, parallelizable regions correspond to sub-expressions of
the program that would be allowed to execute independently by
different processes in the POSIX standard~\cite{posix}. Larger
parallelizable regions can be composed from smaller ones using the
pipe operator (\ttt{|}) and the parallel-composition operator (\ttt{\&}).
Conversely, all other operators, including sequential composition (\ttt{;}) and logical operators (\ttt{\&\&}, \ttt{||}), represent
barrier constructs that do not allow parallelizable regions to expand beyond them.

@@

\heading{Translation Pass}
%
%% \kk{It might be a good idea to change this subsection based on the new
%%   as-late-as-possible architecture where everything is wrapped and
%%   translated at the final moment. Ideally, this would become the main
%%   point of this subsection, as the translation algorithm is kind of
%%   redundant when having the distributable regions too.}
%
\sys's front-end performs a depth-first search on the AST of
the given shell program.  During this pass, it extends the
parallelizable regions bottom-up, translating their
independent components to DFG nodes until a barrier
construct is reached. All AST subtrees not translatable to DFGs are kept as they are. The output of the translation pass is the
original AST where parallelizable regions have been replaced with
DFGs and calls to \sys's runtime primitives~\sx{optimizer}.

@@

\subsection{DFG Definitions}
\label{graph-components}

@@

We first introduce the notation used in this section.
For a set $D$, we write $D\kstar$ to denote the set of all finite words over $D$.
For words $x, y \in D\kstar$, we write $x \cdot y$ or $xy$ to denote their concatenation.
We write $\eps$ for the empty word.
We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there is a word $z$ such that $y = xz$.
The $\leq$ order is reflexive, antisymmetric, and transitive (\ie it is a partial order), and is often called the \emph{prefix order}.

@@

\subsection{Backend: From DFGs to Shell Scripts}
\label{backend}

After translating the parallelizable regions of the input script to DFGs and applying optimization passes~\sx{ir:transformations},
  \sys translates all DFGs back into a shell script.
Nodes of the graph are instantiated with the commands and flags they represent, and edges are instantiated as named pipes.
A prologue in the script creates the necessary intermediate pipes, and a \ttt{trap} call takes care of cleaning things up in cases where the script aborts.
The script makes use of \sys-provided primitives available in the system \ttt{PATH} and described next.

@@

This section describes technical challenges related to execution of the resulting script and how they are addressed by \sys's custom primitives.
%% it instantiates the \todo{custom} split and relay nodes as highly
%% optimized commands to improve In this section we describe how \sys
%% achieves that, and several technical challenges that it addresses.
